The role of prenatal conditions on brain mitochondrial physiology and prey detection in a lizard

### 2025-04-04 ms.qmd
updates

_Metabolic function_
Fresh homogenate suspensions were stained with 5 µL of a fluorescent probe mix containing equal parts 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes as indicators of mitochondrial density, mitochondrial membrane potential, and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample, which we used to distinguish live, viable, intact cells from cellular debris. Within two hours after dissection, these samples were analyzed by flow cytometry.

@@

_Flow Cytometry_
Flow cytometry assays were performed using a Becton Dickson LSRFortessa X-20 flow cytometer with the default wavelength filters on detectors. The detectors and voltage settings used in data acquisition for each were determined during pilot trials and kept consistent throughout the experiment. Data was imported into FlowJo (v. 10.1) for processing. We obtained the mean fluorescent intensity for mitochondrial density, mitochondrial potential, ROS production, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see Methods: flow cytometry in Supplementary Material

#### Statistical analyses
We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, mitochondrial potential, ROS, DNA damage, lipid peroxidation, and detection latency) separately. We first fitted a set of preliminary models for all the interest varaibles (detection latency, mitochondrial density, mitochondrial potential, ROS, DNA damage, and lipid peroxidation) where we included the main effects of hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction . Models also included the sex and age of the lizards at the time where the trials started (for detection latency) or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency, we also included previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation as fixed effects. We included the interaction between CORT and motivation because CORT can impact appetite [@conde2018stress]. All models included clutch identity as a random factor to account for clutch effects. For models of detection latency we also included a random effect of lizard identity because of our repeated measures design. The structure and results of these models are provided on Tables S1-S12 of the Supplementary Material. After the preliminary models, we simplified models by re-fitting to exclude factors that were not significant. However, our models always included the main interest factors: CORT, temperature and their interaction. Random factors remain the same for the final models. All the response variables were mean centered and standardized by dividing by two times the standard deviation [@gelman2008scaling]. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

@@

```{r, fig-results_energy}
#| label: fig-results_energy
#| fig-cap: "Estimates of mitochondrial density (A, C) and mitochondrial potential (B, D) in the olfactory bulbs (A, B) and optic tecta (C, D) of L. delicata hatchlings as a function of the different prenatal conditions. Black dots indicate the posterior mean, and the bars represent the SD of the estimates. The y-axis represents the posterior estimates of the variable of interest, and the x-axis represents the different prenatal conditions. Lines with asterisks represent significant differences between groups based on pMCMC values (pMCMC < 0.05), no lines indicate no significant differences."
#| fig-name: "fig-results_energy"
#
source(here("R", "func.R"))
#
# A) Plotting the results for OB/Chemical stimulus using the df from before
plot_density_OB <- plotting(MD_OB, "Mit density")
plot_potential_OB <- plotting(MP_OB, "Mit potential")
#
fig_OB_energy <- plot_grid(plot_density_OB, NULL, plot_potential_OB, NULL,
                    nrow = 1, rel_widths = c(0.9, 0.1, 0.9, 0.3))
#
# B) Plotting the results for OT/Visual stimulus
plot_density_OT <- plotting(MD_OT, "Mit density")
plot_potential_OT <- plotting(MP_OT, "Mit potential")
#
fig_OT_energy <- plot_grid(plot_density_OT, NULL, plot_potential_OT, NULL,
                    nrow = 1, rel_widths = c(0.9, 0.1, 0.9, 0.3))
#
# Getting the legend (get_legend() does not work if the legend is on the top or bottom)
plot_legend_bottom <- plotting(MP_OB, "Mit potential") + theme(legend.position = "bottom", legend.title = element_blank())
gtable <- ggplot_gtable(ggplot_build(plot_legend_bottom))
legend_plot <- gtable$grobs[[which(sapply(gtable$grobs, function(x) x$name) == "guide-box")]]
# C) Merging plots
#
# Create the figure grid with extra space for images
fig_results_energy <- plot_grid(
  fig_OB_energy, fig_OT_energy, NULL,
  nrow = 3, rel_heights = c(0.9, 0.9, 0.15))
# Final composition: Merge everything, adding images and legend
final_plot_energy <- ggdraw(fig_results_energy) +
  # Insert OB image in the top right
  draw_image("./Others/OB.png", x = 0.87, y = 0.79, width = 0.12, height = 0.2) +
  # Insert OT image in the middle right
  draw_image("./Others/OT.png", x = 0.87, y = 0.34, width = 0.11, height = 0.18) +
  # Insert legend at the bottom-right
  draw_grob(legend_plot, x = 0.45, y = 0.05, width = 0.0001, height = 0.0001) +
  # Insert title for each plot
  annotate("text", x = 0.032, y = 0.97, label = "A", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.487, y = 0.97, label = "B", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.032, y = 0.515, label = "C", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.49, y = 0.515, label = "D", hjust = 1, vjust = 1, size = 7, fontface = "bold")
#
# Print final plot
ggsave(here("./output/figures/text/results_energy.png"), plot = final_plot_energy, width = 21, height = 14, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/text/results_energy.png")
```

_Optic tecta_: we did not find significant effects of CORT, temperature, or their interaction on mitochondrial density, mitochondrial potential, or ROS production in the optic tecta, neither of CORT or incubation temperature on DNA damage or lipid peroxidation (see @fig-results_energy, @fig-results_oxidative, and Table S2). However, the interaction of CORT and temperature have significant effects on DNA damage in the optic tecta (see @fig-results_oxidative). For DNA damage, CORT elevations had no effect on hot-incubated lizards (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Hot - DNA_OT$CORT_Hot), 3, equal = TRUE)`), but it decreased total DNA damage in cold-incubated animals (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Cold - DNA_OT$CORT_Cold), 3, equal = TRUE)`) (see @fig-results_oxidative C, E). Prenatal CORT had opposite effects on eggs incubated at high temperatures (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Hot - LP_OT$CORT_Hot), 3, equal = TRUE)`) than at low temperatures (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Cold - LP_OT$CORT_Cold), 3, equal = TRUE)`), but differences between groups were not significant (see @fig-results_oxidative C). In addition, we found that females had higher DNA damage levels, and lipid peroxidation increased with age (see Table S4).

@@

_Detection of chemical stimulus and its relation with mitochondrial physiology in Olfactory bulbs_: we found no significant effects of mitochondrial function or oxidative stress on prey detection. There were no significant effects between any other of the variables either (see @fig-sem_results_OB and Table S5).

@@

#### Mitochondrial function and ROS production is robust to early life conditions
We found no significant effects of early environmental conditions on mitochondrial density, membrane potential, or ROS production in either of the brain regions studied. This suggests that, contrary to our predictions, mitochondrial function is robust to prenatal temperature and CORT. Previous studies have shown that high incubation temperatures - but not CORT - decreased mitochondrial efficiency in the liver of _L. delicata_ [@crino2024eggs]. Mitochondrial potential is expected to enhance ATP production per oxygen consumed, thereby increasing efficiency [@rice2002brain]. Given the brain's high energy demands [@mcnay_decreases_2000; @mann_coupling_2021], it is plausible that mitochondria are particularly well-protected, which may help explain their resilience towards environmental inputs. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

@@

In this regard, mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. However, the lack of effect of mitochondrial density and mitochondrial potential on detection latency abilities in _L. delicata_ suggests that energy production is not a limiting factor for prey perception, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability. Similarly, the levels of damage in the brain regions studied may not have been high enough to impair prey perception in this species or compensatory mechanisms may have counteracted any negative effects of oxidative stress. However, oxidative damage can accumulate over time, potentially pronouncing the effects of oxidative damage on cognition over time [@terman2006oxidative; @hara_presynaptic_2014]. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Future research should investigate the long-term consequences of mitochondrial activity on brain health and the potential mechanisms that sustain cognitive resilience in _L. delicata_.

#### Conclusions
Prenatal incubation and CORT exposure had region-specific effects in the brain of _L. delicata_. While mitochondrial function remained resilient to early environmental conditions, oxidative damage and behavioral responses exhibited context-dependent variations. Our findings suggest that the brain regions studied are not directly related to the behaviors tested, or that other mechanisms are more relevant for preceiving prey stimuli. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

@@

Assays of oxidative damage from cryopreserved samples were performed within 6 months of the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples by briefly (1-2 minutes) submerging them in near-boiling water. We washed each thawed sample twice, the first time resuspending the pelleted cells in 1 mL warm 1X Tris-EDTA, and the second time in 200 µL warm 1X PBS. Pooled homogenate was divided into 100 µL aliquots to serve as single-color controls. Thawed, resuspended samples were first stained with 20 µL (10 µL for controls) of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor and incubated at 32 °C for 20 minutes. Following staining, we washed cells to prevent further binding of unbound fluorescent probes, then resuspended the pellet in 200 µL warm 1X PBS containing 5 mM (6.14 mg/mL) digitonin. We incubated the samples at 32 °C for 20 minutes to permeabilize the cell membrane, after which we washed the homogenate and resuspended the pelleted cells in 200 µL 1X PBS. We added 20 µL of 70 µM 8-OHdG Polyclonal Antibody to each sample, after which we left the homogenate overnight (~12 hours) for the antibody to bind to 8-OHdG, a marker of oxidative damage on DNA. The following day we counterstained the cells with 20 µL of #µM (100 µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 at 32 °C for 20 minutes. After the cells had been tagged with 8-OHdG antibodies and counterstained, we washed the cells once more and resuspended the pellet in 400 µL of 1X PBS. Unstained and single-color controls were treated identically to samples, but stained with only up to one of BODIPY 665/676, Hoechst 33342, 8-OHdG antibody, or Alexa-Fluor 488 conjugate. Additionally, one control was stained with both 8-OHdG antibody and the Alexa-Fluor 488 conjugated. We then loaded a 96-well plate with 100 µL of each single-color control and 100 µL in duplicate of each sample. We performed all flow cytometry assays on samples within 48-hours of thawing the samples.

@@

Table S5. Compiled direct, indirect, and total effects of SEM for OB/Chemical stimulus
```{r, table_sem_results_OB}
#| label: table_sem_results_OB
#| fig-cap: ""
#
# Table created from the df sem_results_OB (see above)
source(here("R", "func.R"))
#
# Modify database
table_semOB_df <- sem_results_OB %>%
  mutate(`Direct effects` = paste0(mean_direct_effects, " [", q5_direct, ", ", q95_direct, "]"),
         `Indirect effects` = paste0(mean_indirect_effects, " [", q5_indirect, ", ", q95_indirect, "]"),
         `Total effects` = paste0(mean_total_effects, " [", q5_total, ", ", q95_total, "]")) %>%
  dplyr::select(source, predictor_modulator, `Direct effects`, `Indirect effects`, `Total effects`) %>%
  mutate(
      predictor_modulator = case_when(
        predictor_modulator == "mean_mitodensity" ~ "Mitochondrial density",
        predictor_modulator == "mean_potential" ~ "Mitochondrial potential",
        predictor_modulator == "mean_dnadamage" ~ "DNA damage",
        predictor_modulator == "mean_peroxidation" ~ "Lipid peroxidation",
        predictor_modulator == "ROS" ~ "ROS",
        TRUE ~ predictor_modulator
      ),
      source = case_when(
        source == "det_results_semOB" ~ "Detection latency",
        source == "dna_results_semOB" ~ "DNA damage",
        source == "perox_results_semOB" ~ "Lipid peroxidation",
        source == "ros_results_semOB" ~ "ROS",
        TRUE ~ source
      ),
      `Direct effects` = case_when(`Direct effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Direct effects`),
      `Indirect effects` = case_when(`Indirect effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Indirect effects`),
      `Total effects` = case_when(`Total effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Total effects`)) %>%
  rename(Predictor = predictor_modulator, Response = source) %>%
  mutate(Response = factor(Response, levels = c("Detection latency", "DNA damage", "Lipid peroxidation", "ROS")),
         Predictor = factor(Predictor, levels = c("Mitochondrial density",
                                                "Mitochondrial potential",
                                                "ROS",
                                                "DNA damage",
                                                "Lipid peroxidation"))) %>%
  arrange(Response, Predictor)
#
# Table format
set_flextable_defaults(
 font.family = "Times New Roman",
 font.size = 10)
#
table_semOB <- flextable(table_semOB_df) %>%
  align(align = "center", part = "header") %>%
  flextable::compose(i = c(2:5,7,8,10,11,13), j = 1, value = as_paragraph(""), part = "body") %>% # To remove some of the values in the first column
  flextable::hline(i = c(5,8,11), part = "body") %>% 
  autofit()
#
table_semOB
```

@@

Table S9. Summary of the preliminary model for Mitochondrial Potential in Olfactory Bulbs.
```{r, results_preliminary_potential_OB}
#| tbl-cap: "Preliminary results of the models testing for Mitochondrial Potential in OB"
#| label: results_preliminary_potential_OB
#
sum_m_potential_OB_prel <- m_prel_mean_potential_OB %>%
  dplyr::select(-starts_with("r_"), -sigma, -Intercept, -lprior, -lp__) %>%  # Removes random effect terms
  summarise_draws() %>%
  mutate(across(where(is.numeric), ~ as.numeric(format_dec(.x, 3))))
#
flextable(sum_m_potential_OB_prel)
```

@@

Table S10. Summary of the preliminary model for Mitochondrial Potential in Optic Tecta.
```{r, results_preliminary_potential_OT}
#| tbl-cap: "Preliminary results of the models testing for Mitochondrial Potential in OT"
#| label: results_preliminary_potential_OT
#
sum_m_potential_OT_prel <- m_prel_mean_potential_OT %>%
  dplyr::select(-starts_with("r_"), -sigma, -Intercept, -lprior, -lp__) %>%  # Removes random effect terms
  summarise_draws() %>%
  mutate(across(where(is.numeric), ~ as.numeric(format_dec(.x, 3))))
#
flextable(sum_m_potential_OT_prel)
```

@@

```{r , plotmod_potential_OB, out.width="70%", fig.align="center"}
#| label: plotmod_potential_OB
#| caption: "Posterior predictive checks for the model of Mitochondrial Potential in Olfactory Bulbs."
#
mod <- readRDS(here("output/models/mean_potential_def_OB.rds"))
plot_mod <- plot(mod, plot = FALSE)
#
plot_mod_1 <- plot_mod[[1]]
plot_mod_2 <- plot_mod[[2]]
#
fig_mod <- plot_grid(plot_mod_1, plot_mod_2, ncol = 1)
ggsave("./output/figures/suppl/Figure_SX5.png", plot = fig_mod,
       width = 20, height = 28, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/suppl/Figure_SX5.png")
```

Fig S3. Posterior predictive checks for the model of Mitochondrial Potential in Olfactory Bulbs.
Formula: mean_potential ~ cort * temp + (1|clutch)

@@

>>>>

_Metabolic function and ROS_
Fresh homogenate suspensions were stained with 5 µL of a fluorescent probe mix containing equal parts 5 µM MitoTracker Deep Red FM, 2.5 µM MitoTracker Orange CMTMRos, and 50 µM MitoSOX Red. We used these fluorescent probes as indicators of mitochondrial density, mitochondrial membrane potential - a metric of mitochondrial metabolic capacity [@martinez2016tca] - and superoxide (ROS) production, respectively. We also added 5 µL of #µM Hoechst 33342 Nuclear Viability Dye to each sample, which we used to distinguish live, viable, intact cells from cellular debris. Within two hours after dissection, these samples were analyzed by flow cytometry.

@@

_Flow Cytometry_
Flow cytometry assays were performed using a Becton Dickson LSRFortessa X-20 flow cytometer with the default wavelength filters on detectors. The detectors and voltage settings used in data acquisition for each were determined during pilot trials and kept consistent throughout the experiment. Data was imported into FlowJo (v. 10.1) for processing. We obtained the mean fluorescent intensity for mitochondrial density, metabolic capacity, ROS production, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see Methods: flow cytometry in Supplementary Material

#### Statistical analyses
We performed the analyses for each brain region/stimulus and each variable (mitochondrial density, metabolic capacity, ROS, DNA damage, lipid peroxidation, and detection latency) separately. We first fitted a set of preliminary models for all the interest varaibles (detection latency, mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation) where we included the main effects of hormone (CORT versus Control), temperature (Cold versus Hot), and their interaction . Models also included the sex and age of the lizards at the time where the trials started (for detection latency) or when the lizards were euthanized (for all mitochondrial-related variables). For detection latency, we also included previous experience with the prey (familiar versus unknown), their performance on the motivation test [if they ate the cricket (1), or not (0)], and the interaction between CORT and motivation as fixed effects. We included the interaction between CORT and motivation because CORT can impact appetite [@conde2018stress]. All models included clutch identity as a random factor to account for clutch effects. For models of detection latency we also included a random effect of lizard identity because of our repeated measures design. The structure and results of these models are provided on Tables S1-S12 of the Supplementary Material. After the preliminary models, we simplified models by re-fitting to exclude factors that were not significant. However, our models always included the main interest factors: CORT, temperature and their interaction. Random factors remain the same for the final models. All the response variables were mean centered and standardized by dividing by two times the standard deviation [@gelman2008scaling]. Before standardization, mitochondrial density, DNA damage, lipid peroxidation, and detection latency were log-transformed.

@@

```{r, fig-results_energy}
#| label: fig-results_energy
#| fig-cap: "Estimates of mitochondrial density (A, C) and metabolic capacity (B, D) in the olfactory bulbs (A, B) and optic tecta (C, D) of L. delicata hatchlings as a function of the different prenatal conditions. Black dots indicate the posterior mean, and the bars represent the SD of the estimates. The y-axis represents the posterior estimates of the variable of interest, and the x-axis represents the different prenatal conditions. Lines with asterisks represent significant differences between groups based on pMCMC values (pMCMC < 0.05), no lines indicate no significant differences."
#| fig-name: "fig-results_energy"
#
source(here("R", "func.R"))
#
# A) Plotting the results for OB/Chemical stimulus using the df from before
plot_density_OB <- plotting(MD_OB, "Mit density")
plot_potential_OB <- plotting(MP_OB, "Mit potential")
#
fig_OB_energy <- plot_grid(plot_density_OB, NULL, plot_potential_OB, NULL,
                    nrow = 1, rel_widths = c(0.9, 0.1, 0.9, 0.3))
#
# B) Plotting the results for OT/Visual stimulus
plot_density_OT <- plotting(MD_OT, "Mit density")
plot_potential_OT <- plotting(MP_OT, "Mit potential")
#
fig_OT_energy <- plot_grid(plot_density_OT, NULL, plot_potential_OT, NULL,
                    nrow = 1, rel_widths = c(0.9, 0.1, 0.9, 0.3))
#
# Getting the legend (get_legend() does not work if the legend is on the top or bottom)
plot_legend_bottom <- plotting(MP_OB, "Mit potential") + theme(legend.position = "bottom", legend.title = element_blank())
gtable <- ggplot_gtable(ggplot_build(plot_legend_bottom))
legend_plot <- gtable$grobs[[which(sapply(gtable$grobs, function(x) x$name) == "guide-box")]]
# C) Merging plots
#
# Create the figure grid with extra space for images
fig_results_energy <- plot_grid(
  fig_OB_energy, fig_OT_energy, NULL,
  nrow = 3, rel_heights = c(0.9, 0.9, 0.15))
# Final composition: Merge everything, adding images and legend
final_plot_energy <- ggdraw(fig_results_energy) +
  # Insert OB image in the top right
  draw_image("./Others/OB.png", x = 0.87, y = 0.79, width = 0.12, height = 0.2) +
  # Insert OT image in the middle right
  draw_image("./Others/OT.png", x = 0.87, y = 0.34, width = 0.11, height = 0.18) +
  # Insert legend at the bottom-right
  draw_grob(legend_plot, x = 0.45, y = 0.05, width = 0.0001, height = 0.0001) +
  # Insert title for each plot
  annotate("text", x = 0.032, y = 0.97, label = "A", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.487, y = 0.97, label = "B", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.032, y = 0.515, label = "C", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.49, y = 0.515, label = "D", hjust = 1, vjust = 1, size = 7, fontface = "bold")
#
# Print final plot
ggsave(here("./output/figures/text/results_energy.png"), plot = final_plot_energy, width = 21, height = 14, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/text/results_energy.png")
```

_Optic tecta_: we did not find significant effects of CORT, temperature, or their interaction on mitochondrial density, metabolic capacity, or ROS production in the optic tecta, neither of CORT or incubation temperature on DNA damage or lipid peroxidation (see @fig-results_energy, @fig-results_oxidative, and Table S2). However, the interaction of CORT and temperature have significant effects on DNA damage in the optic tecta (see @fig-results_oxidative). For DNA damage, CORT elevations had no effect on hot-incubated lizards (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(DNA_OT$Control_Hot) - mean(DNA_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Hot - DNA_OT$CORT_Hot), 3, equal = TRUE)`), but it decreased total DNA damage in cold-incubated animals (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(DNA_OT$Control_Cold) - mean(DNA_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(DNA_OT$Control_Cold - DNA_OT$CORT_Cold), 3, equal = TRUE)`) (see @fig-results_oxidative C, E). Prenatal CORT had opposite effects on eggs incubated at high temperatures (β~Control-Hot~ - β~CORT-Hot~: mean = `r format_dec(mean(LP_OT$Control_Hot) - mean(LP_OT$CORT_Hot),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Hot - LP_OT$CORT_Hot), 3, equal = TRUE)`) than at low temperatures (β~Control-Cold~ - β~CORT-Cold~: mean = `r format_dec(mean(LP_OT$Control_Cold) - mean(LP_OT$CORT_Cold),3)`, pMCMC `r format_p(pmcmc(LP_OT$Control_Cold - LP_OT$CORT_Cold), 3, equal = TRUE)`), but differences between groups were not significant (see @fig-results_oxidative C). In addition, we found that females had higher DNA damage levels, and lipid peroxidation increased with age (see Table S4).

@@

_Detection of chemical stimulus and its relation with mitochondrial physiology in Olfactory bulbs_: we found a significant effect of mitochondrial density on ROS production (mean β~Mit density - ROS~ = `r format_dec(mean(OB_coeff_mitodensity_ros), 3)`, pMCMC `r format_p(pmcmc(OB_coeff_mitodensity_ros), 3, equal = TRUE)`). However, there were no significant relationships between mitochondrial function or oxidative stress on prey detection, nor between other any other variables (see @fig-sem_results_OB and Table S5).

@@

#### Mitochondrial function and ROS production is robust to early life conditions
We found no significant effects of early environmental conditions on mitochondrial density, membrane potential, or ROS production in either of the brain regions studied. This suggests that, contrary to our predictions, mitochondrial function is robust to prenatal temperature and CORT. Previous studies have shown that high incubation temperatures - but not CORT - decreased mitochondrial efficiency in the liver of _L. delicata_ [@crino2024eggs]. Mitochondrial potential is expected to enhance ATP production per oxygen consumed, thereby increasing metabolic efficiency [@rice2002brain; @martinez2016tca]. Given the brain's high energy demands [@mcnay_decreases_2000; @mann_coupling_2021], it is plausible that mitochondria are particularly well-protected, which may help explain their resilience towards environmental inputs. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

@@

In this regard, mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. However, the lack of effect of mitochondrial density and metabolic capacity on detection latency abilities in _L. delicata_ suggests that energy production is not a limiting factor for prey perception, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability. Similarly, the levels of damage in the brain regions studied may not have been high enough to impair prey perception in this species or compensatory mechanisms may have counteracted any negative effects of oxidative stress. **FOR INSTANCE, ANIMALS MAY UPREGULATE ANTIOXIDANTS FUNCTION WHEN HIGHER RESPIRATION (ONDI'S COMMENT)** 
**Talk about mitochondrial density and its effect on ROS**
Despite not seeing any effect of oxidative damage, since DNA damage or lipid peroxidation can accumulate over time, their effects on cognition can pronounce over time [@terman2006oxidative; @hara_presynaptic_2014]. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Future research should investigate the long-term consequences of mitochondrial activity on brain health and the potential mechanisms that sustain cognitive resilience in _L. delicata_.

#### Conclusions
Prenatal incubation and CORT exposure had region-specific effects in the brain of _L. delicata_. While mitochondrial function remained resilient to early environmental conditions, oxidative damage and behavioral responses exhibited context-dependent variations. Our findings suggest that the brain regions studied are not directly related to the behaviors tested, or that other factors, like motivation, are more relevant for preceiving prey stimuli. Future studies should explore how the environment influences mitochondrial function accross tissues, as well as the mechanisms underlying the brain’s resilience to early-life conditions.

@@

Assays of oxidative damage from cryopreserved samples were performed within 6 months of the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples by briefly (1-2 minutes) submerging them in warm water. We washed each thawed sample twice, the first time resuspending the pelleted cells in 1 mL warm 1X Tris-EDTA, and the second time in 200 µL warm 1X PBS. Pooled homogenate was divided into 100 µL aliquots to serve as single-color controls. Thawed, resuspended samples were first stained with 20 µL (10 µL for controls) of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor and incubated at 32 °C for 20 minutes. Following staining, we washed cells to prevent further binding of unbound fluorescent probes, then resuspended the pellet in 200 µL warm 1X PBS containing 5 mM (6.14 mg/mL) digitonin. We incubated the samples at 32 °C for 20 minutes to permeabilize the cell membrane, after which we washed the homogenate and resuspended the pelleted cells in 200 µL 1X PBS. We added 20 µL of 70 µM 8-OHdG Polyclonal Antibody to each sample, after which we left the homogenate overnight (~12 hours) for the antibody to bind to 8-OHdG, a marker of oxidative damage on DNA. The following day we counterstained the cells with 20 µL of #µM (100 µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 at 32 °C for 20 minutes. After the cells had been tagged with 8-OHdG antibodies and counterstained, we washed the cells once more and resuspended the pellet in 400 µL of 1X PBS. Unstained and single-color controls were treated identically to samples, but stained with only up to one of BODIPY 665/676, Hoechst 33342, 8-OHdG antibody, or Alexa-Fluor 488 conjugate. Additionally, one control was stained with both 8-OHdG antibody and the Alexa-Fluor 488 conjugated. We then loaded a 96-well plate with 100 µL of each single-color control and 100 µL in duplicate of each sample. We performed all flow cytometry assays on samples within 48-hours of thawing the samples.

@@

Table S5. Compiled direct, indirect, and total effects of SEM for OB/Chemical stimulus
```{r, table_sem_results_OB}
#| label: table_sem_results_OB
#| fig-cap: ""
#
# Table created from the df sem_results_OB (see above)
source(here("R", "func.R"))
#
# Modify database
table_semOB_df <- sem_results_OB %>%
  mutate(`Direct effects` = paste0(mean_direct_effects, " [", q5_direct, ", ", q95_direct, "]"),
         `Indirect effects` = paste0(mean_indirect_effects, " [", q5_indirect, ", ", q95_indirect, "]"),
         `Total effects` = paste0(mean_total_effects, " [", q5_total, ", ", q95_total, "]")) %>%
  dplyr::select(source, predictor_modulator, `Direct effects`, `Indirect effects`, `Total effects`) %>%
  mutate(
      predictor_modulator = case_when(
        predictor_modulator == "mean_mitodensity" ~ "Mitochondrial density",
        predictor_modulator == "mean_potential" ~ "Metabolic capacity",
        predictor_modulator == "mean_dnadamage" ~ "DNA damage",
        predictor_modulator == "mean_peroxidation" ~ "Lipid peroxidation",
        predictor_modulator == "ROS" ~ "ROS",
        TRUE ~ predictor_modulator
      ),
      source = case_when(
        source == "det_results_semOB" ~ "Detection latency",
        source == "dna_results_semOB" ~ "DNA damage",
        source == "perox_results_semOB" ~ "Lipid peroxidation",
        source == "ros_results_semOB" ~ "ROS",
        TRUE ~ source
      ),
      `Direct effects` = case_when(`Direct effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Direct effects`),
      `Indirect effects` = case_when(`Indirect effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Indirect effects`),
      `Total effects` = case_when(`Total effects` == "NaN [NA, NA]" ~ " - ", TRUE ~ `Total effects`)) %>%
  rename(Predictor = predictor_modulator, Response = source) %>%
  mutate(Response = factor(Response, levels = c("Detection latency", "DNA damage", "Lipid peroxidation", "ROS")),
         Predictor = factor(Predictor, levels = c("Mitochondrial density",
                                                "Metabolic capacity",
                                                "ROS",
                                                "DNA damage",
                                                "Lipid peroxidation"))) %>%
  arrange(Response, Predictor)
#
# Table format
set_flextable_defaults(
 font.family = "Times New Roman",
 font.size = 10)
#
table_semOB <- flextable(table_semOB_df) %>%
  align(align = "center", part = "header") %>%
  flextable::compose(i = c(2:5,7,8,10,11,13), j = 1, value = as_paragraph(""), part = "body") %>% # To remove some of the values in the first column
  flextable::hline(i = c(5,8,11), part = "body") %>% 
  autofit()
#
table_semOB
```

@@

Table S9. Summary of the preliminary model for Metabolic capacity in Olfactory Bulbs.
```{r, results_preliminary_potential_OB}
#| tbl-cap: "Preliminary results of the models testing for metabolic capacity in OB"
#| label: results_preliminary_potential_OB
#
sum_m_potential_OB_prel <- m_prel_mean_potential_OB %>%
  dplyr::select(-starts_with("r_"), -sigma, -Intercept, -lprior, -lp__) %>%  # Removes random effect terms
  summarise_draws() %>%
  mutate(across(where(is.numeric), ~ as.numeric(format_dec(.x, 3))))
#
flextable(sum_m_potential_OB_prel)
```

@@

Table S10. Summary of the preliminary model for Metabolic capacity in Optic Tecta.
```{r, results_preliminary_potential_OT}
#| tbl-cap: "Preliminary results of the models testing for Metabolic capacity in OT"
#| label: results_preliminary_potential_OT
#
sum_m_potential_OT_prel <- m_prel_mean_potential_OT %>%
  dplyr::select(-starts_with("r_"), -sigma, -Intercept, -lprior, -lp__) %>%  # Removes random effect terms
  summarise_draws() %>%
  mutate(across(where(is.numeric), ~ as.numeric(format_dec(.x, 3))))
#
flextable(sum_m_potential_OT_prel)
```

@@

```{r , plotmod_potential_OB, out.width="70%", fig.align="center"}
#| label: plotmod_potential_OB
#| caption: "Posterior predictive checks for the model of Metabolic capacity in Olfactory Bulbs."
#
mod <- readRDS(here("output/models/mean_potential_def_OB.rds"))
plot_mod <- plot(mod, plot = FALSE)
#
plot_mod_1 <- plot_mod[[1]]
plot_mod_2 <- plot_mod[[2]]
#
fig_mod <- plot_grid(plot_mod_1, plot_mod_2, ncol = 1)
ggsave("./output/figures/suppl/Figure_SX5.png", plot = fig_mod,
       width = 20, height = 28, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/figures/suppl/Figure_SX5.png")
```

Fig S3. Posterior predictive checks for the model of Metabolic capacity in Olfactory Bulbs.
Formula: mean_potential ~ cort * temp + (1|clutch)

@@

#### Brain validation
